---
title: "Homework 05: Factor and figure management"
author: "Stephen Chignell"
date: "October 19, 2018"
output:
  html_document:
    toc: yes
always_allow_html: yes
---

## Overview

**Goals:**

- Reorder a factor in a principled way based on the data and demonstrate the effect in arranged data and in figures.
- Write some data to file and load it back into R.
- Improve a figure (or make one from scratch), using new knowledge, e.g., control the color scheme, use factor levels, smoother mechanics.
- Make a plotly visual.
- Implement visualization design principles.


## Part 1: Factor management

**Overview of factors**

Factors are “truly categorical” variables. They are vectors that:

- have character entries on the surface (i.e. category/class name)
- have integers underneath (i.e. code for computer to keep track)
- have different levels (i.e., number of unique categories/classes)

We will use the gapminder dataset to explore working with factors in the following ways:

- Drop factor / levels
- Reorder levels based on knowledge from data


#### Load libraries

```{r libraries}
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(gapminder))
```

#### Look at the data

```{r gap levels}
str(gapminder)
```

This confirms that we have two factors, `country` (142 levels), and `continent` (5 levels).


#### Drop Oceania

Suppose we wanted to remove observations in the continent of **Oceania** from the dataset. 

To start, we should review the names of the level using the `nlevels` function:

```{r level check}
levels(gapminder$continent)
```

Next we filter the data. We can remove unused levels at the same time by piping the filter results into the `droplevels()` function:

```{r remove Oceania}
gap_sans_OCE <- gapminder %>% 
  filter(continent != "Oceania") %>% 
  droplevels()
```

Now let's confirm there are no longer any "Oceania" rows:
```{r confirm}
NROW(gap_sans_OCE %>% 
  filter(continent == "Oceania"))
```

Looks good! We can also check how many rows were dropped:
```{r difference}
# Subtract to find difference
NROW(gapminder) - NROW(gap_sans_OCE)
```

We have now confirmed that all "Oceania" rows (n=24) have been removed from the original dataset. What about the levels?

```{r check levels}
levels(gap_sans_OCE$continent)
```

Great! We have successfully removed the unused levels.

#### Reorder the levels of `country` or `continent`


Suppose we wanted to plot mean life expectancy for each continent:


```{r mean_pop}
mean_pop <- gapminder %>%
  group_by(continent) %>% 
  summarize(mean.pop = mean(pop)) 
knitr::kable(mean_pop)
```

Let's plot the reordered data:
```{r plot mean_pop}
ggplot(mean_pop, aes(continent, mean.pop)) + 
  geom_point()
```

By default, the levels are ordered alphabetically, which does not help to highlight any patterns in the data. 

We can change these to ascending order using the `fct_reorder` function:

```{r mean_pop ordered}
mean_pop %>%
  mutate(continent = fct_reorder(continent, mean.pop)) %>% 
  ggplot(aes(continent, mean.pop)) + 
  geom_point()
```

## Part 2: File I/O

Let's practice exporting and importing data. Say we wanted to look at life expectancies in American countries with a population over 5,000,000 for the year 2007:

#### Filter and arrange data
```{r filter Americas pop}
America_pop_2007 <- gapminder %>%
  filter(continent == "Americas", year == 2007, pop > 5000000) %>%
  arrange(pop) # arrange in ascending order
```

Let's view it as a table:
```{r view pop table}
knitr::kable(America_pop_2007)
```

#### Write to .CSV
Let's practice exporting to a comma separated values (.csv) file:
```{r write csv}
# set row.names to FALSE to avoide creation of extra ID column
write.csv(America_pop_2007, file = "America_pop_2007_5M.csv", row.names = FALSE) 
```

We can see that the .csv was successfully exported to the project folder and the data are correct in Microsoft Excel. 

Does its structure survive if we re-import it into R?

#### Re-import from .csv
```{r read csv}
America_pop_2007_READ <- read.csv("America_pop_2007_5M.csv")
knitr::kable(America_pop_2007_READ)
```
Great! The data are still arranged by intrecing population, so it surved the write-out/read-in process. 

**Note**: To write or read a .csv outside of the project folder, you need to enter in the full file path.


## Part 3: Visualization design

In this section I reproduce an old figure and give it some new life through the techniques I've learned in the last couple of weeks of lecture. 

#### Old Graph
Below is the code and graph ("as is") for a figure I was very proud of from Homework 02:
```{r old graph}
old_graph <- select(gapminder, gdpPercap, lifeExp, pop, year, continent) %>% #subsetting data
  filter(year > 1957, continent=="Europe"|continent=="Africa") %>% #filter by criteria
  ggplot(aes(lifeExp, gdpPercap))+ #piped to ggplot
  geom_point(aes(color=continent, size=pop, alpha=0.1)) + # add aesthetics 
  xlab("Life Expectancy")+
  ylab("GDP per Capita")
old_graph
```


#### New Graph
```{r new graph}
new_graph <- select(gapminder, gdpPercap, lifeExp, pop, year, continent) %>% 
  filter(year > 1957, continent == "Europe"|continent == "Africa") %>%   ggplot(aes(lifeExp, gdpPercap)) +
  geom_point(aes(color = continent, size = pop, alpha = 0.1)) +
  labs(title = "Life expectancy for each continent", 
       x = "Life Expectancy", 
       y = "GDP per Capita") + 
  theme_bw()+
  theme(axis.text.x  = element_text(vjust=0.6, size=10),
        axis.text = element_text(size = 10)) +
  scale_colour_brewer(palette = "Set1") + # use brewer scheme
  geom_smooth(method="lm") #add a linear regression line
new_graph
```


#### Export to plotly
```{r plotly}
new_graph_plotly <- new_graph
ggplotly(new_graph_plotly) 


```

#### Reflections

One of the first changes I made was to remove the unnecessary comments, as these did not provide extra information and make the code difficult to read, violating the recommendations by Hadley Wickham in the [tidyverse style guide](http://style.tidyverse.org/). Also in line with the style guide, I added spaces before and after each of the "=" and "+" in order to improve legibility. 

I also added a linear regression line, which is particularly nice to include on the plotly plot, as it allows you to identify the predicted values from the model interactively. I could see this--along with the fact that plotly provide access to the specific attribute information for each point--being extremely useful for exploring and presenting data in the future.  


## Part 4: Writing figures to file

With `ggsave()` it's easy to export a figure to as a file on your local computer: 
```{r ggsave}
ggsave("new_graph.png", new_graph)
```

We also may want to adjust the dimensions and resolution:
```{r ggsave dim}
ggsave("new_graph_5x4.png", width = 5, height = 4, dpi = 150)
```

We can also save to vector format with .pdf
```{r ggsave pdf}
ggsave("new_graph.pdf", new_graph)
```

The ggsave( ) function will, by default, save the most recent plot in R. To save a different image, we have to make this explicit in the ggsave code. For example, if we wanted to save the old graph (see above), we would make the following changes to the code:

```{r ggsave old}
ggsave("old_graph.png", plot = old_graph)
```


