---
title: "Homework 05: Factor and figure management"
author: "Stephen Chignell"
date: "October 19, 2018"
output:
  github_document:
    toc: yes
---

## Overview

**Goals:**

- Reorder a factor in a principled way based on the data and demonstrate the effect in arranged data and in figures.
- Write some data to file and load it back into R.
- Improve a figure (or make one from scratch), using new knowledge, e.g., control the color scheme, use factor levels, smoother mechanics.
- Make a plotly visual.
- Implement visualization design principles.


## Part 1: Factor management

**Overview of factors**

Factors are “truly categorical” variables. They are vectors that:

- have character entries on the surface (i.e. category/class name)
- have integers underneath (i.e. code for computer to keep track)
- have different levels (i.e., number of unique categories/classes)

### Exercise using gapminder

We will use the gapminder dataset to explore working with factors in the following ways:

- Drop factor / levels
- Reorder levels based on knowledge from data


#### Load libraries

```{r libraries}
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(gapminder))
```

#### Look at the data

```{r gap levels}
str(gapminder)
```

This confirms that we have two factors, `country` (142 levels), and `continent` (5 levels).


#### Drop Oceania

Suppose we wanted to remove observations in the continent of **Oceania** from the dataset. 

To start, we should review the names of the level using the `nlevels` function:

```{r level check}
levels(gapminder$continent)
```

Next we filter the data:

```{r remove Oceania}
gap_no_Oc <- gapminder %>% 
  filter(continent != "Oceania")
```

And confirm there are no longer any "Oceania" rows:
```{r}
NROW(gap_no_Oc %>% 
  filter(continent == "Oceania"))
```

We can also check how many rows were dropped:
```{r difference}
# Subtract to find difference
NROW(gapminder) - NROW(gap_no_Oc)
```

We have now confirmed that all "Oceania" rows (n=24) have been removed from the original dataset. What about the levels?

```{r check levels}
levels(gap_no_Oc$continent)
```

We still have "Oceania" as an unused level in the new data frame! We can remove this using the `droplevels()` function:






  levels(gap_no_Oc$continent)
```

Ensuring the variable(s) you’re exploring are indeed factors, you are expected to:

    Drop factor / levels;
    Reorder levels based on knowledge from data.



