---
title: "Homework 05: Factor and figure management"
author: "Stephen Chignell"
date: "October 19, 2018"
output:
  github_document:
    toc: yes
---

## Overview

**Goals:**

- Reorder a factor in a principled way based on the data and demonstrate the effect in arranged data and in figures.
- Write some data to file and load it back into R.
- Improve a figure (or make one from scratch), using new knowledge, e.g., control the color scheme, use factor levels, smoother mechanics.
- Make a plotly visual.
- Implement visualization design principles.


## Part 1: Factor management

**Overview of factors**

Factors are “truly categorical” variables. They are vectors that:

- have character entries on the surface (i.e. category/class name)
- have integers underneath (i.e. code for computer to keep track)
- have different levels (i.e., number of unique categories/classes)

## Exercise using gapminder

We will use the gapminder dataset to explore working with factors in the following ways:

- Drop factor / levels
- Reorder levels based on knowledge from data


#### Load libraries

```{r libraries}
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(gapminder))
```

#### Look at the data

```{r gap levels}
str(gapminder)
```

This confirms that we have two factors, `country` (142 levels), and `continent` (5 levels).


#### Drop Oceania

Suppose we wanted to remove observations in the continent of **Oceania** from the dataset. 

To start, we should review the names of the level using the `nlevels` function:

```{r level check}
levels(gapminder$continent)
```

Next we filter the data. We can remove unused levels at the same time by piping the filter results into the `droplevels()` function:

```{r remove Oceania}
gap_sans_OCE <- gapminder %>% 
  filter(continent != "Oceania") %>% 
  droplevels()
```

Now let's confirm there are no longer any "Oceania" rows:
```{r}
NROW(gap_sans_OCE %>% 
  filter(continent == "Oceania"))
```

Looks good! We can also check how many rows were dropped:
```{r difference}
# Subtract to find difference
NROW(gapminder) - NROW(gap_sans_OCE)
```

We have now confirmed that all "Oceania" rows (n=24) have been removed from the original dataset. What about the levels?

```{r check levels}
levels(gap_sans_OCE$continent)
```

Great! We have successfully removed the unused levels.

#### Reorder the levels

If we were to create a plot of the factor `continent`
```{r}
cont <- gap_sans_OCE$continent
```

We see that it is plotted in the order of the factor levels:

```{r, fig.width=4, fig.height=1.5}
qplot(cont)
```

Suppose we wanted to plot life expectancy in Asia. 

```{r}
gap_asia <- gapminder %>% 
  filter(continent == "Asia") %>% 
  droplevels()
```

```{r}
gap_asia %>% 
  group_by(country) %>% 
  summarize(maxlifeExp = max(lifeExp))
```



